Precise spiking motifs in neurobiological and neuromorphic data

### 2022-12-12 manuscript.tex
Update on Overleaf.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction: Importance of precise spike timings in the brain}\label{sec:time}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Is there a neural code?}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% CONTEXT
Neural activity is directly influenced by our immediate environment and by internal states, and is structured in order to generate motor actions. The efficiency of these actions is key for survival, which is the sole determinant in the light of natural selection. A central question of modern neuroscience is to better understand the essence of neural activity, as exemplified by the recordings that can be observed in neurobiological experiments. One sometimes uses the expression ``decoding the neural code'', although this implies the existence of a code, i.e. an explicit representation within the neural activity. Nevertheless, we will use this terminology in all generality to denote the existence of a structure in neural activity. In this respect, it is known that neural activity may be related to specific measurable variables, and since Galvani's experiments, we know that an electrical activity applied on muscular nerves can cause the stretching of a frog's limb (for a review, see~\citep{piccolino_luigi_1997}). A central and well studied way of communication between neurons are specific electrochemical events called action potentials, or spikes, which were first discovered by \href{https://en.wikipedia.org/wiki/Edgar_Adrian}{Lord Adrian}~\citep{adrian_impulses_1926}. In the scope of this article, we focus mainly on these spikes, which are brief (about one millisecond) impulses that propagate along the axons of neurons. First, these have the particular property of being essentially binary in their amplitude, i.e. they are prototypical, 
``\href{https://en.wikipedia.org/wiki/All-or-none_law}{all-or-none}''. Typically, these are received by a neuron from other afferent neurons on the arborized ``input'' dendrite, integrated in its soma, and possibly further propagated along its ``output'' axon to reach efferent neurons. The frequency at which these spikes are emitted has been shown to be roughly commensurate with the stretch of the frog's limb~\citep{adrian_impulses_1926}, and this variable (measured in spikes per second, or Hertz) is as of today considered to be the main constituent of the neural code. Indeed, neurophysiologists typically use the temporal evolution of the firing rate (for instance, as computed as the average occurrence of spikes in small temporal windows of about $100~\ms$) in order to characterize the dynamical activity of neurons. This may be extended by computing different statistics on each neuron's sequence of spikes~\citep{perkel_neuronal_1967} but also the dependence across neurons~\citep{perkel_neuronal_1967-1}.

% GAP
However, computational neuroscience models have suggested that the precise timing within a sequence of spikes may play a crucial role, and that neurons may be as well integrators than synchrony detectors~\citep{abeles_role_1982}. We will investigate this very hypothesis in this review. %Indeed, the response of one biological neuron largely depends on the precise timing of the sequence of incoming spikes that reach its soma~\citep{paugam-moisy_computing_2012}. 
In comparison to a classical analogical vector of inputs, this \emph{event-based representation} observed in the neural code is essential in understanding information processing~\citep{carr_processing_1993}. For instance, it expands the capabilities of such representations (which rely on the firing rate) by considering a representation based on repetitions of spiking motifs at precise times of occurrence. This hypothesis is directly inspired by neurobiological observations and expands the capabilities of analog representations (which are based on spiking firing rates) by considering a representation based on instances of spiking motifs at precise times of occurrence. Additionally, numerous studies demonstrated the importance of precise timing in neural population activity~\citep{davis_spontaneous_2021}, efficient encoding thanks to the use of spike latencies~\citep{perrinet_coding_2004,gollisch_rapid_2008} or precise timing in the auditory system~\citep{deweese_binary_2003,carr_circuit_1990}. All these findings, and more~\citep{bohte_evidence_2004}, highlight the importance of the temporal aspect of the neural code and further suggest the existence of precise spatio-temporal spiking motifs in the input which excites neurons.  A mathematical formalization would be particularly well suited to neuromorphic computing~\citep{roy_towards_2019}, and would allow for the supervised or self-supervised learning of such motifs in any event-driven data.  Crucially, validating this hypothesis would also be crucial in our understanding of neural processes. %
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Dynamics of vision and consequences on the neural code}%Precise timing in the neural code enables efficient vision}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
Let us start with a focus on the state-of-the-art of the role of dynamics in vision. Broadly speaking, vision is the set of processes that allow us to make sense of the luminous world, and is an intensively studied field in neuroscience, in particular with respect to deciphering the neural code. In most mammals, light enters the eye to induce neural activity on the retina, which maintains a certain similitude between the topology of external visual space and its representation on the retina, called retinotopy. The origins of this neuroscientific question can be found in the first experiments from Pierre Flourens which, using lesions in animals, demonstrated the relationship between visual sensations and activity in the cerebral cortex~\citep{flourens_recherches_1842,pearce_marie-jean-pierre_2009}. This was also observed by Lord Adrian when recording the activity the frog's visual system~\citep{adrian_impulses_1926}. In a series of seminal studies,~\citet{hubel_receptive_1968} showed that this activity could be selective to different features, such as the visual orientation or motion of elementary contours. For a large proportion of neurons, there is remarkably a monotonic relationship between the contrast of visual features and the firing frequency of neurons. However, there is no consensus to explain the multiple nonlinear mechanisms that transform the visual scene into retinotopic neural activity maps, even though these processes seem to constitute essential pieces to this puzzle~\citep{carandini_normalization_2012}.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction: Importance of precise spike timings in the brain}\label{sec:time}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Is there a neural code?}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% CONTEXT
Neural activity is directly influenced by our immediate environment and by internal states and is structured in order to generate motor actions. The efficiency of these actions is key for survival, which is the sole determinant in the light of natural selection. A central question of modern neuroscience is to better understand the essence of neural activity, as exemplified by the recordings observed in neurobiological experiments. One sometimes uses the expression ``decoding the neural code'', although this implies the existence of a code, i.e. an explicit representation within the neural activity. Nevertheless, we will use this terminology in all generality to denote the existence of a structure in neural activity. In this respect, it is known that neural activity may be related to specific measurable variables, and since Galvani's experiments, we know that an electrical activity applied on muscular nerves can cause the stretching of a frog's limb (for a review, see~\citep{piccolino_luigi_1997}). A central and well-studied way of communication between neurons is specific electrochemical events called action potentials, or spikes, which were first discovered by \href{https://en.wikipedia.org/wiki/Edgar_Adrian}{Lord Adrian}~\citep{adrian_impulses_1926}. In the scope of this article, we focus mainly on these spikes, which are brief (about one millisecond) impulses that propagate along the axons of neurons. First, these have the particular property of being essentially binary in their amplitude, i.e. they are prototypical, 
``\href{https://en.wikipedia.org/wiki/All-or-none_law}{all-or-none}''. Typically, these are received by a neuron from other afferent neurons on the arborized ``input'' dendrite, integrated in its soma, and possibly further propagated along its ``output'' axon to reach efferent neurons. The frequency at which these spikes are emitted has been shown to be roughly commensurate with the stretch of the frog's limb~\citep{adrian_impulses_1926}, and this variable (measured in spikes per second, or Hertz) is as of today considered to be the main constituent of the neural code. Indeed, most neurophysiologists use the temporal evolution of the firing rate (for instance, as computed as the average occurrence of spikes in small temporal windows of about $100~\ms$) in order to characterize the dynamical activity of neurons. This may be extended by computing different statistics on each neuron's sequence of spikes~\citep{perkel_neuronal_1967} but also the dependence across neurons~\citep{perkel_neuronal_1967-1}.

% GAP
However, computational neuroscience models have suggested that the precise timing within a sequence of spikes may play a crucial role and that neurons may be synchrony detectors as well as integrators~\citep{abeles_role_1982}. We will investigate this very hypothesis in this review. %Indeed, the response of one biological neuron largely depends on the precise timing of the sequence of incoming spikes that reach its soma~\citep{paugam-moisy_computing_2012}. 
In comparison to a classical analog vector of inputs, this \emph{event-based representation} observed in the neural code is essential in understanding information processing~\citep{carr_processing_1993}. For instance, it expands the capabilities of representations, that rely on the firing rate, by considering a representation based on repetitions of spiking motifs at precise times of occurrence. This hypothesis is directly inspired by neurobiological observations and expands the capabilities of analog representations (which are based on spiking firing rates) by considering a representation based on instances of spiking motifs at precise times of occurrence. Additionally, numerous studies demonstrated the importance of precise timing in neural population activity~\citep{davis_spontaneous_2021}, efficient encoding thanks to the use of spike latencies~\citep{perrinet_coding_2004,gollisch_rapid_2008} or precise timing in the auditory system~\citep{deweese_binary_2003,carr_circuit_1990}. All these findings, and more~\citep{bohte_evidence_2004}, highlight the importance of the temporal aspect of the neural code and further suggest the existence of precise spatio-temporal spiking motifs in the input which excites neurons.  A mathematical formalization would be particularly well suited to neuromorphic computing~\citep{roy_towards_2019}, and would allow for the supervised or self-supervised learning of such motifs in any event-driven data.  Crucially, validating this hypothesis would also be crucial in our understanding of neural processes. %
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Dynamics of vision and consequences on the neural code}%Precise timing in the neural code enables efficient vision}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
Let us start with a focus on the state-of-the-art of the role of dynamics in vision. Broadly speaking, vision is the set of processes that allow us to make sense of the world through luminous signals, and is an intensively studied field in neuroscience, in particular with respect to deciphering the neural code. In most mammals, light enters the eye to induce neural activity on the retina, which maintains a certain similitude between the topology of external visual space and its representation on the retina, called retinotopy. The origins of this neuroscientific question can be found in the first experiments from Pierre Flourens which, using lesions in animals, demonstrated the relationship between visual sensations and activity in the cerebral cortex~\citep{flourens_recherches_1842,pearce_marie-jean-pierre_2009}. This was also observed by Lord Adrian when recording the activity the frog's visual system~\citep{adrian_impulses_1926}. In a series of seminal studies,~\citet{hubel_receptive_1968} showed that this activity could be selective to different features, such as the visual orientation or motion of elementary contours. For a large proportion of neurons, there is remarkably a monotonic relationship between the contrast of visual features and the firing frequency of neurons. However, there is no consensus to explain the multiple nonlinear mechanisms that transform the visual scene into retinotopic neural activity maps, even though these processes seem to constitute essential pieces to this puzzle~\citep{carandini_normalization_2012}.
